How to improve your relationship with your future self

### 2016-12-14 responsetoreviewers.md

Hi Cristian and Carsten,

Thanks for giving us the chance to revise the paper on workflow.

In response to your comments and those of the reviewer we revised the paper. Because we realize that the issue is really about transparency and not experiments in particular, we did not end up mentioning experiments in our current set of revisions. We could add a few mentions here and there if that would help.

About the comment that much of the material appeared in an non-peer reviewed journal previously: When I was approached to participate in this journal, I volunteered to send this previously non-peer reviewed article as a submission to be peer-reviewed. That idea was accepted. As I worked to update the previously published version, I invited Maarten Voors to work on the paper with me given shared interests and to diversify the paper (i.e. to ensure that this was not merely the opinion of one single person based on one set of experiences). Maarten has a PhD in economics and works in an economics department, so we both agreed that we would have different perspectives which would enhance the paper. So the version that we sent differs a bit more from the 2011 version than I had anticipated. That said, I don't currently see us writing a whole new paper on this topic. We really appreciate that you allowed our paper to go through the peer-review process and very much hope that the version that we attach here is satisfactory. It is still more of a version 2.0 than a completely new set of thoughts on the topic.

We have attached the file here in pdf format and will happily follow with LaTeX and image sources when needed. Don't hesitate to let us know if you would like other revisions.

@@

Jake and Maarten

Regarding the comments of Reviewer 1:

We very much appreciate the thoughtful and supportive and helpful review. We
respond to the points made here.

 - We have now pointed out explicitly that we have updated and changed the paper to reflect our own changes thinking and the changes in the technology landscape.

@@

 - We have kept the primary lessons the same: we submitted this manuscript to RCP in order to gain more impact from the non-peer reviewed version, and this idea was approved by the folks working on the project at that time. We also feel like the primary lessons stand the test of time even as the specific methods for achieving them might change.

 - The editors highlighted to us that the issue is not primarily about experiments but rather about transparency, so we have not mentioned experiments more in this version.

@@

 - We hope that we have clarified the role of Dropbox, and about makefiles and the make system.

 - We also have edited the paper to try to emphasize that there may be many best practices as long as they all are created to keep in mind the central lessons of the paper. (For example, we believe that one can do transparent scientific workflow using many different programming languages, computing systems, and processes as long as the central lessons are followed.)

 - We have tried to quash the typos and formatting problems although we expect that your copyeditors will help continue this work.

### 2016-12-14 workflow.Rmd
Pre-release version, sent to RCP editors

Memory is tricky. Learning requires effort. When we do not practice and repeat something that we want to remember, most people forget quickly, are overconfident in their abilities to recall future information [@koriat2005illusions], and may even recall events that never happened.^[See the following site for a nice overview of what we know about memory --- including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] Moreover, most of us live busy lives. We type text knowing that the laundry needs doing, hearing children play or cry, ignoring the news or the latest journal review. Our lives and minds are full and we need to efficiently move from task to task. If we cannot count on memory, then how can we do science?

@@

Now, in the future if you wonder how 'that plot on page 10' was created, you
will know: (1) 'that plot' is from a file called `fig1.pdf`and (2)
`fig1.pdf`was created in `makefig1.R`. Any changes to the file in the future
will just require some quick edits of commands already written  (provided R
still exists).^[We use data from @norris2015data throughout this paper.]  Even
if in the future R ceases to exist, you (or someone else) will at least be able
to read the plain text commands and use them to write code in a new favorite
statistical computing language: R scripts are written in [plain
text](http://en.wikipedia.org/wiki/Plain_text), and plain text is a format that
will be around as long as computer programmers write computer programs.

@@

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in LaTeX or
`{#fig:giniprot}` in the [pandoc](http://pandoc.org/) version of markdown built
into [RStudio](https://www.rstudio.com/), we do not need to keep track of the
figure number, nor do extra work when we reorganize the document in response to
reviewer suggestions. Nor do we need a separate `makefig1.R`file or `fig1.pdf`
file.  Tables and other numerical results are also possible to generate within
the source code of a scholarly paper. For example, if we had omitted the
filename in `makebecktable` above, the LaTeX formatted table would have appeared
within this document itself. For Stata users, there now is
[Markdoc](https://github.com/haghish/MarkDoc) which is very similar to R
markdown.

@@

### Step 2
We analyze data in order to explain something about the world to
other scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis, we will do a better job with the data analysis
itself: we will learn as we focus on teaching others about why we did what we
did, and we will avoid errors and save time as we ensure that others (including
our future selves) can retrace our steps. A paper than can be "run" to
reproduce all of its analyses also instills confidence in readers and can more
effectively spur discussion and learning and cumulation of research.

# Meaningful code requires data.

All files containing commands operating on data must refer to a data file. A
reference to a data file is a line of code the analysis program will use to
operate on ('load'/ 'open '/ 'get '/ 'use') the data file. One should not have
to edit this line on different computers or platforms in order to execute this
command.  Using R, for example, all analysis files should have
`load("thedata.rda")`or `read.csv("thedata.csv")`or some equivalent line in
them, and `thedata.csv`should be stored in some easy to find place (like in the
same directory as the file or perhaps in `"Data/thedata.rda"`). Of course, its
even better to include a comment pointing to the data file in addition to the
line loading the file itself.

@@

The principle of modularity suggests that you separate data cleaning,
processing, recoding, and merging from analysis in different files
[@nagler1995coding]. So, perhaps your analysis oriented files will
`load("cleandata.rda")`and a comment in the code will alert the future you
(among others) that `cleandata.rda`was created from `create-cleandata.R` which
in turn begins with
`read.csv(\url("http://www.greatfreedata.gov/dirtydata.csv"))`. Such a data
processing file will typically end with something like
`save("cleandata.rda")`so that we are doubly certain about the provenance of
the data.^[Of course, if you need math or paragraphs to explain what is
happening in these files, you might prefer to make them into R+Markdown or
R+LaTeX files, for which the conventional extension is .Rmd or .Rnw
respectively. So you'd have `create-cleandata.Rnw` written as a mixture of
LaTeX and R which might explain and explore the different coding decisions you
made, perhaps involving some diagnostic plots.]

Now, if in the future we wonder where `cleandata.rda`came from, we
might search for occurrences of `cleandata` in the files on our
system. However, if such searching among files is a burden, an even
nicer solution is to maintain a file for each project called
MANIFEST.txt or INDEX.txt or README.txt which lists the
data and command files with brief descriptions of their functions and
relations.

Best practice is to create a file system where you separate folders by function and separate input from output files. For example, below we show a folder structure for a paper where we look at inequality. The paper is written in .tex, data analysis in Stata (that Maarten prefers, so we see .do and .dta file extensions) in R (the program of choice for Jake, see the .R extensions). Maarten likes to make sure that the directory structure in his projects is the same across projects (so, for example, the numbers on the directory names allow for easy default sorting). The file naming with full dates also allows for easy sorting:

```{bash , eval = FALSE, tidy = FALSE}
00_archive/
01_paper/
  20160622_inequality.tex
02_data/
  00_archive/
  01_rawdata/
  02_cleandata/
  20160622_analysis.dta
  20160622_analysis.rda
03_analysis/
  00_archive/
  01_temp/
  02_output/
    01_tables/
    02_figures/
  20160622_analysis.do
  20160622_analysis.R
  20160622_prep_data.do
  20160622_prep_data.R
 ```

There are a couple of things to note: See that files start with a number to give them an ordering. Note also we create an 00_archive folder, here you can store older files without having to delete them permanently and can reduce the number of files in your main folders. Also there is a clear separation between raw data (data that came straight from the field, downloaded data, someone else's replication files, etc) and clean data (data ready to use). Note that in the analysis folder we created two files, one for cleaning and merging data called `20160622_prep_data` and another for running the analysis called `20160622_analysis`. See also the output folders, with subfolders for graphs and tables. These are place holder for R or STATA to output the results from the analysis and keeps things organized.

Jake tends to be less organized than Maarten and relies on README.md files to tell his future self what is going on in a given directory, a [version control system](http://github.com) to keep track of versions of files, and a [Makefile](http://kbroman.org/minimal_make/) to keep track of relationships between files.

```{bash, tidy = FALSE, eval = FALSE}
manuscript\
figures\
analysis\
  README.md
data\
build\
libraries\
README.md
Makefile
```

**Step 3** We should know where the data came from and what operations were performed on which set of data.

@@

Group work requires version control. Many people are familiar with the 'track
changes' feature in modern WYSIWYG (what you see is what you get) word
processors or the fact that Dropbox allows one to recover previous versions of
files. These are both kinds of version control.  More generally, when we
collaborate, we'd like to do a variety of actions with our shared files.
Collaboration on data analytic projects is more productive and better when (1)
it is easy to see what has changed between versions of files; (2) members of
the team feel free to experiment and then to dump parts of the experimentation
in favor of previous work while merging the successful parts into the main body
of the paper; (3) the team can produce 'releases' of the same document
(one to MPSA, one to APSR, one to their parents) without spawning many possibly
conflicting copies of the same document; (4) people can work on the same files
at the same time without conflicting with one another, and can reconcile their
changes without too much confusion and clobbering. Clobbering is what happens
when your future self or your current collaborator saves an old version of a
file over a new version, erasing good work by accident (or creating a conflicted copy resulting in an arduous back and forth between versions).

Using Dropbox, 'track changes' in Word, or Google Docs is one way to manage your collaborative efforts. They are popular tools, but
have some downsides as they require you to communicate with other folks
in your group before you can edit existing files. On Dropbox, you cannot
edit and save a given file at the same time. Tracking changes with Word has similar limitations (merging the
changes between two documents is not simple or builtin). Google Docs allows for
two people to edit the same document at the same time: but you must be online
with a fairly fast internet connection.^[Note to future selves: change this
section every five years or so. Probably stop saying 'internet' in 2026.] Communication
and trading off editing a file prevents your work (or your colleagues' work)
from getting lost when you both try to save the same file on top of each other. One may also use Dropbox as a kind of server for version
control. See example above. This system is fine untill your collaborations or project grows large
(in terms of MBs) and you may run out of storage space on your computer or you may need to upgrade your Dropbox's storage space). You will however need to agree on an iron rule for file and directory naming with your collaborators to ensure that the files and directories are always named the same across machines and versions (see above).

An excellent, simple, and robust version control system is to rename your files
with the date and time of saving them: yyyymmdd_project.docx (for changes
within the same day, just before a deadline for example, you may even add a
time, so it becomes 20160623_4PM_inequalitypaper.docx). This communicates very
clear what version you are working. Remember the days when you received a file
called inequalityMV4_APSRsubmit_reallyfinal2.tex? That should quickly become
something of the past! We find ourselves preaching the gospel to our
collaborators frequently and will keep renaming files until a collaborator has
been successfully pacified. Be sure to include year in the file names ---
remember, the life of an idea is measured in years. If you are wise enough to
have saved your documents as plain text then you can easily compare documents
using the many utilities available for comparing text files.^[Adobe Acrobat
allows one to compare differences in pdf files. OpenOffice supports a 'Compare
Documents' option. Word now does the same. And Google Docs will report on the
version history of a document.] When you reach certain milestones you can
rename the file accordingly: thedocAPSA2009.tex --- for the one sent to
discussants at APSA --- or thedocAPSR2015.tex --- for the version eventually
sent to the APSR six years after you presented it at APSA. The formal version
control systems mentioned above all allow this kind of thing and are much more
elegant and capable, but you can do it by hand too as long as you don't mind
taking up a lot of disk space and having many 'thedoc... ' files around.
Indeed, the number of files can add up quickly (especially around submission
time!) and you may want to create an "0_Archive" folder to store older
versions. If you do version control by hand, spend a little extra time to
ensure that you do not clobber files when you make mistakes typing in the
file-names. And, if you find yourself spending extra time reconciling changes
made by different collaborators by hand, remember this is a task that modern
version control systems take care of quickly and easily.

However, best practice we feel is to use formal version control approaches to organizing work [version
control](https://en.wikipedia.org/wiki/Version_control) strategy. As of 2016,
the standard for managing large collaborations with many files is the
[Git](https://en.wikipedia.org/wiki/Git_(software)) system. User friendly free
front-ends for Git currently make the process of learning and using Git very
convenient and integrated into the kind of workflow that your future self will
be proud of: [Open Science Framework](http://osf.io)
[bitbucket](http://bitbucket.com), [github](http://github.com). Jake uses
[github](http://github.com) for all of his collaborations with others as well
as with his future self: his collaborators appreciate some of the nice extra
features of GitHub, such as the ability to keep a [shared task
list](https://guides.github.com/features/issues/). Learning about version
control systems takes a bit of time. We suggest, however, it is well worth the
time investment as it will save (lots!) of time later on.

**Step 5** Writing is rewriting. Thus, all writing involves versions. When we
collaborate with ourselves and others we want to avoid clobbering and we want
to enable graceful reconciliation of rewriting. One can do these things with
formal systems of software (like git) or with formal systems of file naming,
file comparing, and communication or, even better, with both. In either case,
plain text files will make such tasks easier, will take up less disk space, and
be easier to read for the future you.

@@

Anyone writing custom code should worry about getting it right. The more code
one writes, the more time one has to appreciate problems arising from bugs,
errors, and typos in data analysis and code. Since mistakes will always be
made, we should just recognize we need more standardization of practice to
minimize the mistakes and help others find them. One thing we can do to
minimize and catch the inevitable mistakes is to include testing in our coding.

Of course, this idea, is not new. The desire to avoid error
looms large when large groups of programmers write code for multi-million
dollar programs. The idea of [test driven
development](http://en.wikipedia.org/wiki/Test-driven_development) and the idea
that one ought to create tests of [small parts of one's
code](http://en.wikipedia.org/wiki/Unit_testing) arose to address such
concerns.^[There are now R packages to help R package writers do this, see for
example <https://github.com/hadley/testthat> and the article on it
<https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf>.]

@@

##  The actual function is:
times.2.fn <- function(thenumber){
	##  This function multiplies a scalar number by 2
	##  thenumber is a scalar number
	thenumber+2
}
```
Here we use the test function to make sure it works.

```{r results = 'asis'}
test.times.2.fn()
```

Ack! we mistyped '+' for '*'. Good thing we wrote the
test!^[A more common example of this kind of testing occurs
everyday when we recode variables into new forms but look at a
crosstab of the old vs. new variable before proceeding.]

@@

**Step 6** No one can forsee all of the ways that a
computer program can fail. One can, however, at least make sure that
it succeeds in doing the task motivating the writing of the code in
the first place.

@@

Over the past decades more and more attention is being paid to 'reproducible research', creating a
'reproducible workflow' and 'literate programming' these days. As always, the devil is in the details: Here we list a few of our own
attempts at enabling reproducible research. You'll find many other inspiring
examples on the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering their
files online for copying and improvement. By the way, if you do copy and
improve, it is polite to alert the person from whom you made the copy about
your work and to credit them in some way.

We have experimented with several systems so far:

(1) We wrote this paper in the [R markdown](http://rmarkdown.rstudio.com/)
literate programming format using two different text editors (Jake used vim)
and (Maarten used [atom](https://atom.io/)). We organized our files and
collaboration on GitHub. The source code for this document can be accessed,
downloaded and modified from <https://github.com/jwbowers/workflow>. We also
used GitHub Issues to send notes to each other and maintain a task list. The
nice thing about GitHub is that is enables you to make "releases" of a project
which enable you to smooth the reproduction of all of the products of your
research (simulations, data analyses, tables, figures, etc..).  Similar
features are offered by the [Open Science Framework (OSF)](https://osf.io/).
Maarten recently participated in a workshop on research transparency organized
by the [Berkeley Institute for Transparency in the Social
Sciences](www.bitss.org) in which all [course material](https://osf.io/qh2nr/)
was accessible through OSF.  OSF integrates with Github nicely and a key
benefit is that both systems have been created to remain for a long time
(unlike perhaps your university drive).  (2) For one paper Jake simply mixed R
and LaTeX code in one document (called the "Sweave" format of literate
programming) and then added that document and data into into a compressed
archive [@bowers2005dataverse]; (3) For another more computing intensive paper
Jake assembled a set of files that enabled reproduction of results using the
`make` system [@bowers2008dataverse]; (4) Jake also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; The benefit
of the compendium approach is that one is not required to have access to a
command line for `make`: the compendium is downloadable from within R using
`install.packages()` and is viewable using the `vignette()` function in any
operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R
packages as creative and energetic folks turn their attention to the question
of reproducible research.^[See for example <http://r-pkgs.had.co.nz/>] Finally,
(5) Maarten has been using Dropbox for most of his collaborative projects. See
comments above. As a tool for making files publicly available Dropbox is less
useful. You can of course make replication files available through creating an
public folder, but you will still need to refer to this folder on a particular
project website (an example is [Maarten's project on conflict and football](
https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl = 0)).

>>>>

Memory is tricky. Learning requires effort. When we do not practice and repeat something that we want to remember, most people forget quickly, are overconfident in their abilities to recall future information [@koriat2005illusions], and may even recall events that never happened.^[See the following site for a nice overview of what we know about memory --- including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] Moreover, most of us live busy lives. We type text knowing that the laundry needs doing, hearing children play or cry, ignoring the news or the latest journal review, worrying about a friend. Our lives and minds are full and we need to efficiently move from task to task. If we cannot count on memory, then how can we do science?

@@

Now, in the future if you wonder how 'that plot on page 10' was created, you
will know: (1) 'that plot' is from a file called `fig1.pdf` and (2)
`fig1.pdf` was created in `makefig1.R`. Any changes to the file in the future
will just require some quick edits of commands already written  (provided R
still exists).^[We use data from @norris2015data throughout this paper.]  Even
if in the future R ceases to exist, you (or someone else) will at least be able
to read the plain text commands and use them to write code in a new favorite
statistical computing language: R scripts are written in [plain
text](http://en.wikipedia.org/wiki/Plain_text), and plain text is a format that
will be around as long as computer programmers write computer programs.

@@

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in LaTeX or
`{#fig:giniprot}` in the [pandoc](http://pandoc.org/) version of markdown built
into [RStudio](https://www.rstudio.com/), we do not need to keep track of the
figure number, nor do extra work when we reorganize the document in response to
reviewer suggestions. Nor do we need a separate `makefig1.R` file or `fig1.pdf`
file.  Tables and other numerical results are also possible to generate within
the source code of a scholarly paper. For example, if we had omitted the
filename in `makebecktable` above, the LaTeX formatted table would have appeared
within this document itself. For Stata users, there now is
[Markdoc](https://github.com/haghish/MarkDoc) which is very similar to R
markdown.

@@

### Step 2
We analyze data in order to explain something about the world to
other scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis, we will do a better job with the data analysis
itself: we will learn as we focus on teaching others about why we did what we
did, and we will avoid errors and save time as we ensure that others (including
our future selves) can retrace our steps. A paper than can be "run" to
reproduce all of analyses also instills confidence in readers and can more
effectively spur discussion and learning and cumulation of research.

# The territory of a data analysis requires maps.

Data analysis tends to involve multiple people working with multiple files.
Future collaborators will need a map to understand the flow of inputs (like raw
data) and outputs (like figures, tables, and individual numbers).

## Meaningful code requires data.

All files containing commands operating on data must refer to a data file. A
reference to a data file is a line of code that the analysis program will use
to operate on ('load' / 'open' / 'get' / 'use') the data file. One should not
have to edit this line on different computers or platforms in order to execute
this command.  Using R, for example, all analysis files should have
`load("thedata.rda")` or `read.csv("thedata.csv")` or some equivalent line in
them, and `thedata.csv` should be stored in some easy to find place (like in
the same directory as the file or perhaps in `"Data/thedata.rda"`). Of course,
its even better to include a comment pointing to the data file in addition to
the line loading the file itself. This means one should not see lines like
`setwd(C:\JakesFiles\theproject)` or `setwd(/Users/maartenvoors/theproject)` at
the beginning of documents that are meant to be shared with a future self or
present or future others.

## File organization can be a map itself

@@

The principle of modularity suggests that you separate data cleaning,
processing, recoding, and merging from analysis in different files
[@nagler1995coding]. So, perhaps your analysis files will
`load("cleandata.rda")` and a comment in the code will alert the future you
(among others) that `cleandata.rda` was created from `create-cleandata.R` which
in turn begins with
`read.csv(\url("http://data.gov/dirtydata.csv"))`. Such a data
processing file will typically end with something like
`save("cleandata.rda")` so that we are doubly certain about the provenance of
the data.^[Of course, if you need math or paragraphs to explain what is
happening in these files, you might prefer to make them into R+Markdown or
R+LaTeX files, for which the conventional extension is .Rmd or .Rnw
respectively. So you'd have `create-cleandata.Rmd` written as a mixture of
markdown and R which might explain and explore the different coding decisions you
made, perhaps involving some diagnostic plots.]

Now, if in the future we wonder where `cleandata.rda` came from, we might
search for occurrences of `cleandata` in the files on our system. Of course, if
it is difficult to find the right version of `cleandata` on the system, it
might help to know where files like `cleandata` tend to be: that is, to have a
system for project file organization.  Best practice is to create a file system
where you separate folders by function and separate input from output files.
For example, below we show a folder structure for a paper where we look at
inequality. The paper is written in .tex, data analysis in both Stata (that
Maarten prefers, so we see .do and .dta file extensions) and in R (the program
of choice for Jake, see the .R extensions). Maarten likes to make sure that the
directory structure in his projects is the same across projects (so, for
example, the numbers on the directory names allow for easy default sorting).
The file naming with full dates also allows for easy sorting:

```{bash , eval = FALSE, tidy = FALSE}
00_archive/
01_paper/
  20160622_inequality.tex
02_data/
  00_archive/
  01_rawdata/
  02_cleandata/
  20160622_analysis.dta
  20160622_analysis.rda
03_analysis/
  00_archive/
  01_temp/
  02_output/
    01_tables/
    02_figures/
      20160622_models.do
      20160612_simulations.R
      20160502_prep_data.do
 ```

There are a couple of things to note: See that files start with a number to give them an ordering. Note also we create an 00_archive folder, here you can store older files without having to delete them permanently and can reduce the number of files in your main folders. Also there is a clear separation between raw data (data that came straight from the field, downloaded data, someone else's replication files, etc) and clean data (data ready to use). Note that in the analysis folder we created two files, one for cleaning and merging data called `20160622_prep_data` and another for running the analysis called `20160622_models`. See also the placeholder output folders, with subfolders for figures and tables.

## The input-output map can also be in the form of files

Another solution to the problem of finding files and knowing how files relate,
is to maintain a file for each project called MANIFEST.txt or INDEX.txt or
README.txt which lists the data and command files with brief descriptions of
their functions and relations. However, this file may be a burden to maintain.

Jake tends to be less organized than Maarten and relies on README.md files to
tell his future self what is going on in a given directory, a [version control
system](http://github.com) to keep track of versions of files (so that he
doesn't need to add dates to filenames), and a
[Makefile](http://kbroman.org/minimal_make/) to keep track of relationships
between files. Notice that the `analysis` subdirectory has its own README.md
file.

```{bash, tidy = FALSE, eval = FALSE}
manuscript\
  paper.pdf
  paper.Rmd
figures\
  boxplots.R
  boxplots.pdf
tables\
analysis\
  README.md
  models.R
data\
  workingdata.R
  rawdata.csv
  workingdata.rda
build\
libraries\
README.md
Makefile
```

```{r echo=FALSE}
## from https://stackoverflow.com/questions/20409172/how-to-display-verbatim-inline-r-code-with-backticks-using-rmarkdown
rinline <- function(code) {
  sprintf('``` `r %s` ```', code)
}
```

Imagine that the `paper.Rmd` shows the `boxplots.pdf` file and also reports on the total size of the data. So, the introduction to the paper requires the `workingdata.rda` file (to know the sample size) and `boxplots.pdf`. In turn, those files depend on `workingdata.R` which takes `rawdata.csv` and produces the `workingdata.rda` file, and `boxplots.R` which uses `workingdata.rda` to make the plots. A Makefile records this web of relationships in a structured way such that one can say `make manuscript/paper.pdf` and all of the files that are required to produce `paper.pdf` will be executed if they have not recently been used or if they have recently been changed.^[See <http://kbroman.org/minimal_make/> for more about Makefiles.] Here is an example that records the relationships just described:

```{bash eval=FALSE}
manuscript/paper.pdf: manuscript/paper.Rmd figures/boxplot.pdf
    cd manuscript && Rscript -e "library(rmarkdown); render('paper.Rmd')"

figures/boxplot.pdf: figures/boxplot.R data/workingdata.rda
    R CMD BATCH figures/boxplot.R

data/workingdata.rda: rawdata.csv workingdata.R
    R CMD BATCH data/workingdata.R
```

To produce `paper.pdf` having only downloaded `rawdata.csv` and the command
files, one would type `make paper.pdf` and the GNU make system would first run
` R CMD BATCH data/workingdata.R` then, if that is successful, it would run `R
CMD BATCH figures/boxplot.R` and finally run the line required for creation of
`paper.pdf`. Later, if only `paper.Rmd` is edited, `make paper.pdf` would only
run the line for `paper.pdf` because it would know that `boxplot.pdf` and
`workingdata.rda` are relatively recent and the .csv and .R files required for
them have not changed.

### Step 3
We should know where the data came from and what operations were performed on which set of data. The authors of this paper have two different versions of this workflow. Both fulfill their purpose as a map to the territory of a data analysis.

@@

Group work requires knowing which versions of files are new, which are old, and
what changed in between. Many people are familiar with the 'track changes'
feature in modern WYSIWYG (what you see is what you get) word processors or the
fact that Dropbox or Google Docs allow one to recover previous versions of
files. These are both kinds of version control.  More generally, when we
collaborate, we'd like to do a variety of actions with our shared files.
Collaboration on data analytic projects is more productive and better when (1)
it is easy to see what has changed between versions of files; (2) members of
the team feel free to experiment and then to dump parts of the experimentation
in favor of previous work while merging the successful parts into the main body
of the paper; (3) the team can produce 'releases' of the same document (one to
MPSA, one to APSR, one to their parents) without spawning many possibly
conflicting copies of the same document; (4) people can work on the same files
at the same time without conflicting with one another, and can reconcile their
changes without too much confusion and clobbering. Clobbering is what happens
when your future self or your current collaborator saves an old version of a
file over a new version, erasing good work by accident (or creating a
conflicted copy resulting in an arduous back and forth between versions).

Using Dropbox, 'track changes' in Word, or Google Docs is one way to manage
your collaborative efforts. They are popular tools, but have some downsides as
they require you to communicate with other folks in your group before you can
edit existing files. On Dropbox, you cannot edit and save a given file at the
same time. Tracking changes with Word has similar limitations (merging the
changes between two documents is not simple or built-in). Google Docs allows
for two people to edit the same document at the same time: but you must be
online with a fairly fast internet connection.^[Note to future selves: change
this section every five years or so. Probably stop saying 'internet' in 2026.]
Communication and trading off while editing a file prevents your work (or your
colleagues' work) from getting lost when you both try to save the same file on
top of each other. One may also use Dropbox as a kind of server for version
control. See example above. This system is fine until your collaborations or
project grows large (in terms of MBs) and you may run out of storage space on
your computer or you may need to upgrade your Dropbox's storage space. You will
however need to agree on an iron rule for file and directory naming with your
collaborators to ensure that the files and directories are always named the
same across machines and versions plus a clear communication plan for trading off.

An excellent, simple, and robust version control system is to rename your files
with the date and time of saving them: yyyymmdd_project.docx (for changes
within the same day, just before a deadline for example, you may even add a
time, so it becomes 20160623_4PM_inequalitypaper.docx). This communicates
version and prevents clobbering unless the old file has the same name or you
don't update the new file name. Remember the days when you received a file
called inequalityMV4_APSRsubmit_reallyfinal2.tex? That should quickly become
something of the past. We find ourselves preaching this gospel to our
collaborators frequently and will keep renaming files until a collaborator has
been successfully pacified. Be sure to include year in the file names ---
remember, the life of an idea is measured in years. If you are wise enough to
have saved your documents as plain text then you can easily compare versions of
the same document using the many utilities available for comparing text
files.^[Adobe Acrobat allows one to compare differences in pdf files.
OpenOffice supports a 'Compare Documents' option. Word now does the same. And
Google Docs will report on the version history of a document. On a Mac the
FileMerge utility works for plain text files.] When you reach certain
milestones you can rename the file accordingly: thedocAPSA2009.tex --- for the
one sent to discussants at APSA --- or thedocAPSR2015.tex --- for the version
eventually sent to the APSR six years after you presented it at APSA. The
formal version control systems mentioned above all allow this kind of thing and
are much more elegant and capable, but you can do it by hand as long as you
don't mind taking up disk space and having many 'thedoc... ' files
around.  Indeed, the number of files can add up quickly (especially around
submission time!) and you may want to create an "0_Archive" folder to store
older versions. If you do version control by hand, spend a little extra time to
ensure that you do not clobber files when you make mistakes typing in the
file-names. And, if you find yourself spending extra time reconciling changes
made by different collaborators by hand, remember this is a task that modern
version control systems take care of quickly and easily.

We feel the best practice is to use formal version control approaches to
organizing work [version
control](https://en.wikipedia.org/wiki/Version_control). As of 2016, the
standard for managing large collaborations with many files is the
[Git](https://en.wikipedia.org/wiki/Git_(software)) system. User friendly free
front-ends for Git currently make the process of learning and using Git very
convenient and integrated into the kind of workflow that your future self will
be proud of:  [github](http://github.com) and [bitbucket](http://bitbucket.com)
make it easier to use git, and the [Open Science Framework](http://osf.io)
integrates with github to further ease this task. Jake uses
[github](http://github.com) for all of his collaborations with others as well
as with his future self: his collaborators appreciate some of the nice extra
features of GitHub, such as the ability to keep a [shared task
list](https://guides.github.com/features/issues/). Learning about version
control systems takes a bit of time. We suggest, however, it is well worth the
time investment as it will save lots of time later on.

### Step 4
Writing is rewriting. Thus, all writing involves versions. When we
collaborate with ourselves and others we want to avoid clobbering and we want
to enable graceful reconciliation of rewriting. One can do these things with
formal systems of software (like git) or with formal systems of file naming,
file comparing, and communication or, even better, with both. In either case,
plain text files will make such tasks easier, will take up less disk space, and
be easier to read for the future you.

@@

Anyone writing custom code should worry about getting it right. The more code
one writes, the more time one has to appreciate problems arising from bugs,
errors, and typos in data analysis and code. One thing we can do to
minimize and catch the inevitable mistakes is to include testing in our coding.

This idea is not new. The desire to avoid error looms large when large groups
of programmers write code for multi-million dollar programs. The idea of [test
driven development](http://en.wikipedia.org/wiki/Test-driven_development) and
the idea that one ought to create tests of [small parts of one's
code](http://en.wikipedia.org/wiki/Unit_testing) arose to address such
concerns.^[There are now R packages to help R package writers do this, see for
example <https://github.com/hadley/testthat> and the article on it
<https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf>.]

@@

##  The actual function is:
times.2.fn <- function(thenumber){
	##  This function multiplies a scalar number by 2
	##  thenumber is a scalar number
	thenumber+2
}
```

Here we use the test function to make sure it works.

```{r}
test.times.2.fn()
```

Ack! we mistyped $+$ instead of $*$. Good thing we wrote the
test!^[A more common example of this kind of testing occurs
everyday when we recode variables into new forms but look at a
crosstab of the old vs. new variable before proceeding.]

@@

### Step 5
No one can forsee all of the ways that a computer program can fail.
One can, however, at least make sure that it succeeds in doing the task
motivating the writing of the code in the first place or produces an error in
the right place and the right time.

@@

Over the past decades more and more attention is being paid to 'reproducible
research', and creating a 'reproducible workflow'. As always, the devil is in
the details: Here we list a few of our own attempts at enabling reproducible
research. You'll find many other inspiring examples on the web. Luckily, the
open source ethos aligns nicely with academic incentives, so we are beginning
to find more and more people offering their files and ideas about workflow
online for copying and improvement. By the way, if you do copy and improve, it
is polite to alert the person from whom you made the copy and to credit them in
some way if not also to offer your ideas for improvement.^[For a formal example
of copying and improving code, see the GitHub based fork and pull-request
workflow <https://gist.github.com/Chaser324/ce0505fbed06b947d962>.]

We have experimented with several systems so far: (1) We wrote this paper in
the [R markdown](http://rmarkdown.rstudio.com/) literate programming format
using two different text editors (Jake used vim) and (Maarten used
[atom](https://atom.io/)). We organized our files and collaboration on GitHub.
The source code for this document can be accessed, downloaded and modified from
<https://github.com/jwbowers/workflow>. We also used GitHub Issues to send
notes to each other and maintain a task list. The nice thing about GitHub is
that is enables you to make "releases" of a project which enable you to smooth
the reproduction of all of the products of your research (simulations, data
analyses, tables, figures, etc..).  Similar features are offered by the [Open
Science Framework (OSF)](https://osf.io/).  Maarten recently participated in a
workshop on research transparency organized by the [Berkeley Institute for
Transparency in the Social Sciences](www.bitss.org) in which all [course
material](https://osf.io/qh2nr/) was accessible through OSF.  OSF integrates
with Github nicely and a key benefit is that both systems have been created to
remain for a long time (unlike perhaps your university drive).  (2) For one
paper Jake simply mixed R and LaTeX code in one document (called the "Sweave"
format of literate programming) and then added that document and data into into
a compressed archive [@bowers2005dataverse]; (3) For another more computing
intensive paper Jake's collaborator, Mark Fredrickson, assembled a set of files
that enabled reproduction of results using the `make` system
[@bowers2008dataverse]; (4) Jake also tried the "compendium" approach
[@gentleman2005reproducible,@gentleman2007statistical] which embeds an academic
paper within the R package system [@bowers2011dataverse]; The benefit of the
compendium approach is that one is not required to have access to a command
line for `make`: the compendium is downloadable from within R using
`install.packages()` and is viewable using the `vignette()` function in any
operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R
packages as creative and energetic folks turn their attention to the question
of reproducible research.^[See for example <http://r-pkgs.had.co.nz/>] Finally,
(5) Maarten has been using Dropbox for most of his collaborative projects. See
comments above. As a tool for making files publicly available Dropbox is less
useful. You can of course make replication files available through creating an
public folder, but you will still need to refer to this folder on a particular
project website (an example is [Maarten's project on conflict and football](
https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl = 0)).
